The Astropy Project: Building an open-science project and status of the v2.0 core package

### 2018-03-01 main.tex
Removing provisional packages section
Change wording as suggested by @crawfordsm

The \astropy project aims to provide an open-source and open development core
package (\astropypkg) and an ecosystem of \emph{affiliated packages} that
support astronomical functionality in the \python programming language.
``Open development'' describes a process where anybody with an internet connection can suggest changes to the source code and contribute their opinion when new features, bug fixes, or other code changes, governance, or any other aspect of the development process is discussed (see Sect.~\ref{sect:devmod} and \ref{sect:ape} of how this is organized in practice).
The \astropypkg core package is now a feature-rich library of sufficiently
general tools and classes that supports the development of more specialized
code. An example of such functionality is reading and writing FITS files: It would be
time consuming and impractical for multiple groups to implement the FITS
standard \citep{FITS} and maintain software for such a general-purpose need.
Another example of such a common task is in dealing with representations of and
transformations between astronomical coordinate systems.

The \astropy project aims to develop and provide high-quality code and
documentation according to the best practices in software development.
The project makes use of different tools and web services to reach those
goals without central institutional oversight.
The first public release of the \astropypkg package is described in
\cite{astropy}. Since then, the \astropypkg package has been
used in hundreds of projects and the scope of the package has grown
considerably. At the same time, the scientific community
contributing to the project has grown tremendously and an ecosystem
of packages supporting or affiliated with the \astropypkg core has
developed.
In this paper, we describe the current status of the \astropy community and the
\astropypkg core package and discuss goals for future development.

We start by describing the way the \astropy project functions and is organized
in \sectionname~\ref{sec:org}. We then describe the main software efforts
developed by the \astropy project itself: a core package called \astropypkg
(\sectionname~\ref{sec:core}) and several separate packages that help maintain
the infrastructure for testing and documentation
(\sectionname~\ref{sec:infrastructure}). We end with a short vision for
the future of \astropy and astronomical software in general in
\sectionname~\ref{sec:future}. The full paper, including the code
to produce the figures, is available in a \github
repository\footnote{\url{https://github.com/astropy/astropy-v2.0-paper}}

@@

The \astropy project, as of the version 2.0 release, does not receive any direct
financial support for the development of \astropypkg.
Development of the software, all supporting materials, and community support are
provided by individuals who work on the \astropy project in their own personal
time, by individuals or groups contributing to \astropy as part of a research
project, or contributions from institutions that allocate people to work on
\astropy.
A list of organizations that have contributed to \astropy in this manner
can be found in the Acknowledgments.

Different funding models have been proposed for support of \astropy
(e.g., \citealt{2016arXiv161003159M}), but a long-term plan
for sustainability has not yet been established.
The \astropy project has the ability to accept financial contributions
from institutions or individuals through the NumFOCUS\footnote{NumFOCUS
is a 501(c)(3) nonprofit that supports and promotes world-class, innovative,
open source scientific computing.} organization. NumFOCUS has, to date, covered the direct costs incurred by the \astropy project.

\section{Astropy Core Package Version 2.0}
\label{sec:core}
The \astropy project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing
astronomical files, and units) or base classes for other
packages to utilize for a common interface (e.g., \texttt{NDData}).
In this section, we highlight new features introduced or substantially improved
since version 0.2 (previously described in \citealt{astropy}).  The \astropypkg
package
provides a full log of changes\footnote{\url{https://github.com/astropy/astropy/blob/stable/CHANGES.rst}}
over the course of the entire project and more details about individual
subpackages are available in the documentation.\footnote{\url{http://docs.astropy.org/en/stable/}}
Beyond what is mentioned below, most subpackages have seen improved performance
since the release of the version 0.2 package.

@@

The \package{astropy.table} subpackage provides functionality for
representing and manipulating heterogeneous data. In some respects,
this is similar to \package{numpy} record arrays \citep{numpy} or
\package{pandas} dataframes \citep{pandas} but with modifications for
astronomical data. Most notably, tables from \package{astropy.table}
allow for table or column metadata and can handle
vectors or arrays as table entries.
The subpackage was described in
detail in \cite{astropy}.  Thus, in what follows, we only summarize
key new features or updates to \package{astropy.table} since the
previous \astropy paper. These are support for grouped table
operations, table concatenation, and using array-valued
\package{astropy} objects as table columns.

@@

>>>>

The \astropy Project aims to provide an open-source and open-development core
package (\astropypkg) and an ecosystem of \emph{affiliated packages} that
support astronomical functionality in the \python programming language.
The \astropypkg core package is now a feature-rich library of sufficiently
general tools and classes that supports the development of more specialized
code. An example of such functionality is reading and writing FITS files: It would be
time consuming and impractical for multiple groups to implement the FITS
standard \citep{FITS} and maintain software for such a general-purpose need.
Another example of such a common task is in dealing with representations of and
transformations between astronomical coordinate systems.

The \astropy Project aims to develop and provide high-quality code and
documentation according to the best practices in software development.
The project makes use of different tools and web services to reach those
goals without central institutional oversight.
The first public release of the \astropypkg package is described in
\cite{astropy}. Since then, the \astropypkg package has been
used in hundreds of projects and the scope of the package has grown
considerably. At the same time, the scientific community
contributing to the project has grown tremendously and an ecosystem
of packages supporting or affiliated with the \astropypkg core has
developed.
In this paper, we describe the current status of the \astropy community and the
\astropypkg core package and discuss goals for future development.

We start by describing the way the \astropy Project functions and is organized
in \sectionname~\ref{sec:org}. We then describe the main software efforts
developed by the \astropy Project itself: a core package called \astropypkg
(\sectionname~\ref{sec:core}) and several separate packages that help maintain
the infrastructure for testing and documentation
(\sectionname~\ref{sec:infrastructure}). We end with a short vision for
the future of \astropy and astronomical software in general in
\sectionname~\ref{sec:future}. The full paper, including the code
to produce the figures, is available in a \github
repository\footnote{\url{https://github.com/astropy/astropy-v2.0-paper}}

@@

The \astropy Project, as of the version 2.0 release, does not receive any direct
financial support for the development of \astropypkg.
Development of the software, all supporting materials, and community support are
provided by individuals who work on the \astropy Project in their own personal
time, by individuals or groups contributing to \astropy as part of a research
project, or contributions from institutions that allocate people to work on
\astropy.
A list of organizations that have contributed to \astropy in this manner
can be found in the Acknowledgments.

Different funding models have been proposed for support of \astropy
(e.g., \citealt{2016arXiv161003159M}), but a long-term plan
for sustainability has not yet been established.
The \astropy Project has the ability to accept financial contributions
from institutions or individuals through the NumFOCUS\footnote{NumFOCUS
is a 501(c)(3) nonprofit that supports and promotes world-class, innovative,
open source scientific computing.} organization. NumFOCUS has, to date, covered the direct costs incurred by the \astropy Project.

\subsection{Reuse of the scientific Python ecosystem}

The \astropy Project is built on a philosophy of building from the existing
Python scientific ecosystem wherever possible. Many of the enabling technologies
like \texttt{numpy}, \texttt{scipy}, \texttt{matplotlib}, or \texttt{cython},
are necessary as the core underpinnings of \astropy packages.  Often this means
using these packages as ``building blocks'' (e.g., the ubiquitous
use of \texttt{numpy} arrays throughout \astropypkg).  In other cases this means
wrappers around more general algorithms that make them more convenient for
astronomy use cases (e.g., the model-fitting in the \package{astropy.modeling}
subpackage discussed in \sectionname~\ref{sec:modeling}).  Sometimes these
simple wrappers evolve into more complex implementations that address
astronomically-relevant use cases the general tool does not support (e.g.,
\package{astropy.convolution}, see \sectionname~\ref{ssec:convolution}). As a
broad rule, the Project explicitly encourages re-use of code where possible.

However, the boundaries of when this re-use is called for often ambiguous.  Some
of the examples above only evolved after significant debate in the community
over whether these algorithms were sufficient.  Other times, even apparently
general functionality did not exist in the wider ecosystem that met the \astropy
community's needs, and hence the functionality had to be developed wholly from
the developer resources available in the community (e.g.,
\package{astropy.units}, or \package{astropy.table} when it began). At the same 
time, however, the \astropy Project has provided the wider community with myriad bug
fixes to the enabling technologies listed above, as well as the testing and
documentation architecture. Functionality is only ``extracted'' from \astropy to 
other packages after careful consideration that includes considering the impact 
on the maintenance and support of \astropy.

\section{Astropy Core Package Version 2.0}
\label{sec:core}
The \astropy Project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing
astronomical files, and units) or base classes for other
packages to utilize for a common interface (e.g., \texttt{NDData}).
In this section, we highlight new features introduced or substantially improved
since version 0.2 (previously described in \citealt{astropy}).  The \astropypkg
package
provides a full log of changes\footnote{\url{https://github.com/astropy/astropy/blob/stable/CHANGES.rst}}
over the course of the entire project and more details about individual
subpackages are available in the documentation.\footnote{\url{http://docs.astropy.org/en/stable/}}
Beyond what is mentioned below, most subpackages have seen improved performance
since the release of the version 0.2 package.

@@

The \package{astropy.table} subpackage provides functionality for
representing and manipulating heterogeneous data. In some respects,
this is similar to \package{numpy} record arrays \citep{numpy} or
\package{pandas} \texttt{DataFrame} objects \citep{pandas}, but with modifications for
astronomical data. Most notably, tables from \package{astropy.table}
allow for table or column metadata and can handle
vectors/arrays or arbitrary objects (with suitable column-like characteristics)
as table entries. The subpackage was described in
detail in \cite{astropy}.  Thus, in what follows, we only summarize
key new features or updates to \package{astropy.table} since the
previous \astropy paper. These are support for grouped table
operations, table concatenation, and using array-valued
\package{astropy} objects as table columns.

@@

Additionally, there is now interoperability between \texttt{Table}
and \package{pandas} \texttt{DataFrame} objects using the \texttt{Table.to_pandas()}
and \texttt{Table.from_pandas()} methods. \texttt{Table.to_pandas()} is,
however, limited to a subset of possible \texttt{Table} objects, because several
of the features outlined above cannot be represented as \package{pandas} data
frames. While these features might be contributed to or combined with
\package{pandas} at a later date, the architectural differences that
enable these areas are significant, so this would be a major undertaking that would 
require substantial investment from the core maintainers and communities of both 
packages.
